The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2019-06-19 razzo-introduction.tex
Folded TeX

Modern computational techniques allow to infer phylogenies from genetic data such as DNA, RNA or proteins. BEAST (\cite{beast}) and its descendant BEAST2 (\cite{beast2}) are widely used tools to perform this task, which they can achieve by running a Bayesian analysis given data and tree priors.

BEAST2 gives to the user the option to set up several possible phylogenetic priors (e.g. substitution/clock/speciation models). However, currently available priors can be not suitable to analyze some specific datasets.

For this reason BEAST2 provides users with the possibility to introduce new tree priors, to infer phylogenies based on different assumptions on how the speciation process takes place.

Current phylogenetic tools assume that only a single speciation event can occur at any given time.
While this assumption has been proved to be useful to construct a wide variety of successful models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{etienne2012diversity}, \cite{etienne2014estimating}), they disallow for environmental changes that trigger speciations in multiple clades at a same point in time.

We explore such case introducing the multiple birth model, currently absent in BEAST. The multiple birth hypothesis aims to include species pump mechanisms to investigate drivers and modes of such diversification processes whose phylogenies show an impressive amount of speciation events in relatively short times.

The (constant-rate) birth-death (BD) model embodies the common assumption that only a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model relaxes this assumption allowing, in addition to standard BD events, also events in which large-scale environmental changes lead to speciation bursts.
Such hypothesis can be useful to better understand the history of systems of particular interests for evolutionary biologists, such as the  diversification of cichlid fish in the African Great Lakes (Malawi, Tanganyika and Victoria), where water level fluctuations are thought to play an important role in promoting diversification (\citep{verheyen1996mitochondrial}, \citep{sturmbauer2001lake}, \citep{janzen2016}, \citep{janzen2017}).

However, the introduction of new tree priors is not always desirable (\citep{bilderbeek2019pirouette}). Current BD tree priors might, in principle, prove to be good enough at detecting such events despite the lower level of complexity. If this is the case one should always be more keen to adopt the simplest model.

We used the R package \verb;pirouette; (\citep{pirouette}) to perform such test, starting on phylogenies simulated under the MBD regime using the \verb;mbd; package (\citep{mbd}).
From such phylogenies we measure the inference error made adopting a standard BD tree prior in the inference process.

With this work we aim, using such inference error distributions, to test whether or not it is advantageous to implement a new prior model that can allow the construction of trees where multiple speciations can co-occur at the same time.

>>>>

Modern computational techniques allow to infer phylogenies from genetic data suc
h as DNA, RNA or proteins. BEAST (\cite{beast}) and its descendant BEAST2 (\cite
{beast2}) are widely used tools to perform this task, which they can achieve by 
running a Bayesian analysis given data and tree priors.

BEAST2 gives to the user the option to set up several possible phylogenetic prio
rs (e.g. substitution/clock/speciation models). However, currently available pri
ors can be not suitable to analyze some specific datasets.

For this reason BEAST2 provides users with the possibility to introduce new tree
 priors, to infer phylogenies based on different assumptions on how the speciati
on process takes place.

Current phylogenetic tools assume that only a single speciation event can occur 
at any given time.
While this assumption has been proved to be useful to construct a wide variety o
f successful models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, \cite{eti
enne2012diversity}, \cite{etienne2014estimating}), they disallow for environment
al changes that trigger speciations in multiple clades at a same point in time.

We explore such case introducing the multiple birth model, currently absent in B
EAST. The multiple birth hypothesis aims to include species pump mechanisms to i
nvestigate drivers and modes of such diversification processes whose phylogenies
 show an impressive amount of speciation events in relatively short times.

The (constant-rate) birth-death (BD) model embodies the common assumption that o
nly a single speciation event can occur at any given time.
The multiple-birth-death (MBD) model relaxes this assumption allowing, in additi
on to standard BD events, also events in which large-scale environmental changes
 lead to speciation bursts.
Such hypothesis can be useful to better understand the history of systems of par
ticular interests for evolutionary biologists, such as the  diversification of c
ichlid fish in the African Great Lakes (Malawi, Tanganyika and Victoria), where 
water level fluctuations are thought to play an important role in promoting dive
rsification (\citep{verheyen1996mitochondrial}, \citep{sturmbauer2001lake}, \cit
ep{janzen2016}, \citep{janzen2017}).

However, the introduction of new tree priors is not always desirable (\citep{bil
derbeek2019pirouette}). Current BD tree priors might, in principle, prove to be 
good enough at detecting such events despite the lower level of complexity. If t
his is the case one should always be more keen to adopt the simplest model.

We used the R package \verb;pirouette; (\citep{pirouette}) to perform such test,
 starting on phylogenies simulated under the MBD regime using the \verb;mbd; pac
kage (\citep{mbd}).
From such phylogenies we measure the inference error made adopting a standard BD
 tree prior in the inference process.

With this work we aim, using such inference error distributions, to test whether
 or not it is advantageous to implement a new prior model that can allow the con
struction of trees where multiple speciations can co-occur at the same time.
